Thermal reactionomes reveal adaptive responses to thermal extremes in warm

### 2014-10-10 doc/ApTranscriptome_manuscript.Rmd
edits to text, mostly Discussion

---
title: Thermal reactionomes reveal adaptive responses to thermal extremes in warm
  and cold-climate ant species
author: ' '
csl: ecology.csl
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: no
    toc_depth: 2
  html_document: null
  word_document: default
includes: header.tex
bibliography: ApTranscriptome.bib
---

@@

John Stanton-Geddes \textsuperscript{1}, Andrew Nguyen \textsuperscript{1}, James Vincent \textsuperscript{2}, Mahesh Vangala \textsuperscript{2}, Robert R. Dunn \textsuperscript{3}, Aaron M. Ellison \textsuperscript{4}, Nathan J. Sanders \textsuperscript{5}, Nicholas J. Gotelli \textsuperscript{1}, Sara Helms Cahan \textsuperscript{1}

@@

\textsuperscript{1} Department of Biology, University of Vermont, Burlington, Vermont 05405 \newline
\textsuperscript{2} Vermont Genetics Network, University of Vermont, Burlington, Vermont 05405 \newline
\textsuperscript{3} Department of Biological Sciences, North Carolina State University, Raleigh, North Carolina 27695 \newline
\textsuperscript{4} Harvard Forest, Harvard University, Petersham, Massachusetts 01336 \newline
\textsuperscript{5} Center for Macroecology, Evolution and Climate, University of Copenhagen

@@

Temperature regulates biological activity and shapes diversity from molecular to macroecological scales [@brown2004; @kingsolver2009]. Many species, especially small insects, live at temperatures close to their thermal limits and are at risk from current increases in temperature [@deutsch2008; @kingsolver2013; @sunday2014]. Thermal tolerance, the ability of individuals to maintain function and survive thermal extremes, depends on a complex interplay between the structural integrity of cellular components and activiation of physiological response mechanisms to prevent and/or repair damage [@huey1989; @richter2010]. Thermal defense strategies are shaped by the environmental regime organisms experience [@angiletta2003] and thermal limits vary considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013]. These differences in thermal tolerance are largely genetic [@krebs1997; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011]. Outside of candidate genes [@krebs1998], little is known about the molecular basis of this variation in thermal tolerance or the link between short-term physiological acclimation and longer-term adaptation to novel temperature regimes. This information is critical for understanding the adaptive potential of species to future climates [@hoffmann2008].

@@

Here, we extend the reaction norm approach to RNAseq analysis with the *reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across an environmental gradient. We use this method to identify genes that are thermally responsive in two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A. picea* [@umphrey1996]. *Aphaenogaster* are some of the most common ant in eastern North America [@king2013] where they are keystone seed dispersers [@ness2009; @zelikova2011; @rodriguez-cabal2012]. *Aphaenogaster carolinensis* (mean annual temperature at collection site = ...) experiences significantly warmer and less seasonally variable temperatures than *A. picea* (mean annual temperature from collection site = ; [@hijmans2005]), with correspondingly higher minimal and maximal critical thermal limits that allow us to investigate adaptation to both lower and upper thermal extremes (...CTmin and CTmax from HF/DF...) [@diamond2012].

To characterize the thermal reactionome, we collected whole organism RNAseq data at 12 temperatures (every 3.5°C between 0 and 38.5°C) using Illumina short read sequencing technology, generated the *de novo* transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis (Supplemental Material). We used these data to quantitatively test three mechanistic hypotheses of thermal adaptation. First, the *enhanced defense* hypothesis (...ref...) proposes that the stress-adapted species extend their thermal limits through a stronger induced response to provide greater protection from damage. This hypothesis would predict that *A. picea* induces more genes in response to low temperatures, whereas *A. carolinensis* induces more genes in response to high temperatures. Additionally, upregulation of molecular pathways shared by both species should be stronger or occur in response to less initial temperatures stress for more commonly-experienced thermal extremes; low temperatures for *A. picea*, high temperatures for *A. carolinensis*. Second, the *genetic assimiliation* hypothesis [@waddington1953; @sikkink2014] proposes that exposure to more frequent stressors selects for a shift from inducible to constitutive expression of stress-response genes. This hypothesis predicts that transcripts responsive to high temperatures in *A. picea* will have higher constitutive expression in *A. carolinensis*, while transcripts responsive to low temperatures in *A. carolinensis* will have higher constitutive expression in *A. picea*. Finally, the *tolerance* hypothesis (...ref...fields 2001?...) proposes that existing inducible stress responses become insufficient or prohibitively costly as environmental stressors increase in frequency, resulting in a shift away from induced response in favor of structural [@fields2001] or behavioral [@gordon2013; @sunday2014] adapations. This hypothesis predicts adaptation to stress should be associated with lower transcriptional responsiveness and less sensitivity to temperature perturbation, as well as a shift to an alternate suite of tolerance genes and pathways. Of course, these hypotheses are not mutually exclusive. By taking a reactionome approach, we can quantify the relative importance of each of these mechanisms to thermal adapatation as the proportion of the transcriptome matching each of these predicted patterns.

@@

Species persistence in face of climate change requires both short-term physiological acclimation and longer-term adapative changes. In this study, we explored the relationship between these processes using whole organism gene expression data from closely-related ant species.

Under the *enhanced defense* hypothesis, we predicted that the cool-climate species *A. picea* would induce more genes in response to low temperatures than *A. carolinensis*, while *A. carolinensis* would induce more genes in response to high temperatures. This prediction was only true for *A. picea*, which induced the same genes and then recruited additional genes in response to both low and high temperatures (Fig. 1). In addition, we predicted that upregulation should occur in response to less extreme low and high temperatures for *A. picea* and *A. carolinensis*, respectively. In contrast, we found no difference between species in the mean temperature of expression response for low temperatures (Fig. 2A,C), and that higher temperatures were required to induce the expression response for *A. carolinensis* than *A. picea* for low temperatures. While conflicting with the *enhanced defense* hypothesis, this is consistent with the *tolerance* hypothesis for *A. carolinensis*.

We also explored the *genetic assimilation* hypothesis that exposure to more frequent stressors would select for a shift from inducible to constitutive expression of stress-response genes. In contrast to this prediction, constitutive expression levels at 19.5°C were higher in *A. carolinensis* than *A. picea* in nearly all comparisons. The exception to this was for genes that had **High** or **Bimodal** expression in *A. picea* and were not responsive in *A. carolinensis*. Rather than being constitutively elevated as predicted, these genes were not expressed (...mean TPM...) in *A. carolinensis*. ...gene ontology terms...genes recruited to thermal tolerance... Consistent with our results but at a much shorter time-scale, no evidence for genetic assimilation at the expression level was found in an experimental evolution study with *Caenorhabditis remanei* selected for 10 generations to increased heat tolerance, even though there was a substantial increase in heat tolerance and loss of plasticity in the evolved populations [@sikkink2014]. In this study, the genetic assimilation at the phenotypic level was caused by a shift in the threshold for temperature response, similar to what we observed here (Fig. 2).

While our results did not the genetic assimilation hypothesis, the observation of higher constitutive expression of thermally-responsive transcripts is again consistent with the *tolerance* hypothesis for *A. carolinensis*.

Thus, we find evidence that these two species have responded to temperature using distinct mechanisms. *A. picea* has enhanced its thermal response, while *A. carolinensis* tolerates thermal stress...

----------------------------------------------------

with the prediction that the warm-climate species *A. carolinensis* would rely more upon "frontloading" than the cool-climate *A. picea*. Moreover, *A. carolinensis* shut-down expression of more transcripts (Fig. 2) at thermal extremes, and had greater inertia of expression to changes in temperature (Fig. 2) than did *A. picea*. In constrast, the cool-climate species *A. picea* activated more transcripts in response to both thermal extremes, including enrichment of biological processes related to metabolism, transport and regulation. One possible explanation for this difference in responses is that *A. picea* mounts a classical cellular stress response [@kultz2005; @richter2010], requiring increased metabolism, while *A. carolinensis* relies upon constitutive defenses [@barshis2013], proteome stability [@ghosh2010] or behavior [@sunday2014] to tolerate thermal stress. Consistent with this hypothesis, we found that of the 7 thermally responsive heat shock proteins, all 7 increased expression in response to thermal extremes in *A. picea*, but only 3 do in *A. carolinensis*. In contrast, by shutting-down molecular activity, must rely upon other mechanisms such as behavior or increased proteome stability to tolerate thermal stress. These differences are in line with the expectations if *A. picea*, a northern species with fewer good foraging days, must forage in suboptimal conditions whereas *A. carolinensis* can, with a longer season, afford to restrain from foraging in suboptimal conditions. In fact, restrained foraging under stressful conditions by the red harvester ant *Pogonomyrmex barbatus* has been shown to increase colony fitness [@gordon2013].

Recent RNA-seq studies of thermal tolerance, using only a few temperature conditions, have identified a few dozen to hundred genes that are differentially expressed between conditions [@meyer2011; @barshis2013; @palumbi2014; @oneil2014]. By applying a regression approach, we were able to quantitatively describe changes in transcript expression across a thermal gradient. In particular, by characterizing the thermal reactionomes of the ant species *Aphaenogaster picea* and *A. carolinensis* we found differences indicative of adaptation to climate, even while the total number of differentially-expressed genes was similar between the two species (Table 1). In total, we found that about 8% of transcripts (~9,000) are thermally responsive in each species, and that about 70% of these transcripts have a non-linear response to temperature that differs between species, which would not have been detected by standard differential expression approaches. Generally, this indicates that the molecular response to thermal extremes is much more complex than indicated by previous studies, as would be expected given the ecological importance of thermal tolerance [@kingsolver2009].

Following the genetic assimilation hypothesis that adaptation occurs via changes from plastically induced to constitutive expression, we predicted a tradeoff between constitutive expression levels and inducibility of a transcript.

**put heat shock protein in Discussion**

> We examined the reaction norms for transcripts in the GO category "response to stress" which we *a priori* expected to have significant thermal responses (Fig. 3). Although `r resp.stress.responsive.count` transcripts were annotated with this GO term among the thermally responsive trancripts, this number was not enriched compared to the total number in the transcriptome ($\chi^2$ = `r round(GO.stress.Xsq$statistic, 2)`, P < `r round(GO.stress.Xsq$p.value, 2)`). In addition, of the `r nrow(hsp_all)` genes annotated to heat shock proteins in the transcriptome, `r length(which(hsp_responsive$A22.type %in% c("Bimodal", "High", "Low")))` were upregulated to low or high temperatures in *A. picea* while only `r length(which(hsp_responsive$Ar.type %in% c("Bimodal", "High", "Low")))` were upregulated in *A. carolinensis*.

@@

CTmax and CTmin hypothesized to be genetically correlated, but not evident in terms of gene expression.  Only 15-20% of genes upregulated in response to temp are bimodal (ie generalized stress response), and although upper activation thresholds differed significantly btwn species, lower thresholds were virtually identical.  Implications . . .

A major contribution of this study is the reaction norm approach for gene expression data (but see Hodgins-Davis et al [-@hodgins-davis2012] for a similar approach using differential expression among many levels). A number of issues have hampered RNA-seq studies to date. Namely, lists of differentially expressed genes are prone to false positives [@gonzalez2013], depend on the genetic background of the organism [@sarup2011], identify genes that are correlated, but not necessarily causative, with phenotypes [@feder2005] and are prone to "storytelling" [@pavlidis2012]. Our findings are robust to these issues as we focus on the average change in the shape of the reaction norms across many hundreds of responsive genes in each species. Thus, while we use gene ontology information to explain our main results, the key findings about the number of responsive genes and shifts in temperatures of induction do not depend on functional annotation. Similarly, in the plant and animal breeding literature there has been a shift from using genome-wide association methods which are prone to the same problems listed above, to using genomic selection methods which provide a direct means to predict a population's response to selection [@goddard2007].

@@

In this work, we have brought reaction norms to the genomic era by characterizing the thermal reactionomes of two temperate ant species, *Aphaenogaster picea* and *A. carolinensis*. At least 10% of their transcriptomes are thermally responsive. For both species, more genes responded to low than high temperatures. Together with physiological data showing more variation in critical minimum than maximum temperatures, characterization of the reactionome suggests that these species, and likely other ectotherms, have less adaptive potential in response to high than low temperatures. Our results also indicate that these two ant species have different responses to thermal extremes. *A. picea* responds by increasing metabolism and activating heat shock proteins and other protective molecules, whereas *A. carolinensis* shuts down metabolism and must rely on other mechanisms for thermal tolerance. The thermal reactionomes of these two species provide key insights into the genetic basis of thermal tolerance, and a resource for the future study of ecological adaptation in ant species. Finally, the reactionome itself illustrates a new direction for characterizing acclimation and adaptation in a changing climate.

@@

To describe the overall molecular toolkit for thermal tolerance in *Aphaenogaster*, we performed gene set enrichment analysis (GSEA) on the intersection of each expression category in the two species using the package `topGO` [@alexa2006; @alexa2010] with the `parentChild` algorithm [@grossmann2007] implemented in R [@r_core_team2013]. Briefly, this approach identifies GO terms that are overrepresented in the significant transcripts relative to all GO terms in the transcriptome, accounting for dependencies among the GO terms. We clustered enriched GO terms from the `topGO` analysis by semantic similarity using the package `GOSemSim` [@yu2010] to reduce redundancy.

@@

The raw Illumina reads are available at [...], the script for analysis and version history is available at (http://github.com/johnstantongeddes/ApTranscriptome).

>>>>

---
title: Thermal reactionomes reveal adaptive responses to thermal extremes in warm
  and cool-climate ant species
author: ' '
csl: ecology.csl
output:
  word_document: default
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: no
    toc_depth: 2
  html_document: null
includes: header.tex
bibliography: ApTranscriptome.bib
---

@@

John Stanton-Geddes \textsuperscript{1}, Andrew Nguyen \textsuperscript{1}, Lacy Chick \textsuperscript{2}, James Vincent \textsuperscript{3}, Mahesh Vangala \textsuperscript{3}, Robert R. Dunn \textsuperscript{4}, Aaron M. Ellison \textsuperscript{5}, Nathan J. Sanders \textsuperscript{2,6}, Nicholas J. Gotelli \textsuperscript{1}, Sara Helms Cahan \textsuperscript{1}

@@

\textsuperscript{1} Department of Biology, University of Vermont, Burlington, Vermont 05405 \newline
\textsuperscript{2} Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, Tenessee 37996 \newline
\textsuperscript{3} Vermont Genetics Network, University of Vermont, Burlington, Vermont 05405 \newline
\textsuperscript{4} Department of Biological Sciences, North Carolina State University, Raleigh, North Carolina 27695 \newline
\textsuperscript{5} Harvard Forest, Harvard University, Petersham, Massachusetts 01336 \newline
\textsuperscript{6} Center for Macroecology, Evolution and Climate, University of Copenhagen

@@

Temperature regulates biological activity and shapes diversity from molecular to macroecological scales [@brown2004; @kingsolver2009]. Many species, especially small insects, live at temperatures close to their thermal limits and are at risk from current increases in temperature [@deutsch2008; @kingsolver2013; @sunday2014]. Thermal tolerance, the ability of individuals to maintain function and survive thermal extremes, depends on a complex interplay between the structural integrity of cellular components and activiation of physiological response mechanisms to prevent and/or repair damage [@huey1989; @richter2010]. Thermal defense strategies are shaped by the environmental regime organisms experience [@angilletta_jr2003] and thermal limits vary considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013]. These differences in thermal tolerance are largely genetic [@krebs1997; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011]. Outside of candidate genes [@krebs1998], little is known about the evolution of thermal tolerance or the link between short-term physiological acclimation and longer-term adaptation to novel temperature regimes. This information is critical for understanding the adaptive potential of species to future climates [@hoffmann2008].

@@

Here, we extend the reaction norm approach to RNAseq analysis with the *reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across an environmental gradient. We use this method to identify genes that are thermally responsive in two closely related eastern North American ant species, *Aphaenogaster carolinensis* (part of the *A. rudis*-species complex, B. DeMarco personal communication) and *A. picea* [@umphrey1996]. *Aphaenogaster* are some of the most common ant in eastern North America [@king2013] where they are keystone seed dispersers [@ness2009; @zelikova2011; @rodriguez-cabal2012]. *Aphaenogaster carolinensis* (mean annual temperature (MAT) at collection site = 14.6°C, seasonality = 7,678) experiences significantly warmer and less seasonally variable temperatures than *A. picea* (MAT = 4.6°C, seasonality = 10,008; [@hijmans2005]), with correspondingly higher critical maximum temperatures (44.7°C for *A. carolinensis* versus 41.4°C for *A. picea*; Supplemental Material) and minimum temperatures (6.1°C for *A. carolinensis* versus -0.1°C for *A. picea*; Supplemental Material) allowing us to investigate adaptation to both lower and upper thermal extremes in this system.

```{r climate, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library(raster)
w <- getData('worldclim', var='bio', res=10)
#names(w)
# Plot first bioclim dimension (MAT) to check download worked
#plot(w, 1, col=colorRampPalette(c("blue","green","red"))(255))
# sites
sites <- data.frame(site = c("Ac", "Ap"), lon=c(-78.874, -72.702), lat= c(36.037, 44.508))
# extract bioclimatic variables from worldclim data. Note bioclim temperature data is multiplied by 10 (no decimal), so transform by dividing by 10.
dbio <- extract(w, sites[,c("lon","lat")])
# bio1 = MAT
# bio4 = Temperature Seasonality (standard deviation *100)
dbio.vars <- cbind(sites, dbio[,1], dbio[,4])
colnames(dbio.vars)[which(colnames(dbio.vars) == "dbio[, 1]")] <- "MATx10"
dbio.vars$MAT <- dbio.vars$MATx10/10
colnames(dbio.vars)[which(colnames(dbio.vars) == "dbio[, 4]")] <- "Seasonality"
dbio.vars
```

To characterize the thermal reactionome, we collected whole organism RNAseq data at 12 temperatures (every 3.5°C between 0 and 38.5°C) using Illumina short read sequencing technology, generated the *de novo* transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis (Supplemental Material). We used these data to quantitatively test three mechanistic hypotheses of thermal adaptation. First, the *enhanced defense* hypothesis [@hofmann1996; @feder1999; @kultz2005] proposes that the stress-adapted species extend their thermal limits through a stronger induced response to provide greater protection from damage. This hypothesis would predict that *A. picea* induces more genes in response to low temperatures, whereas *A. carolinensis* induces more genes in response to high temperatures. Additionally, upregulation of molecular pathways shared by both species should be stronger or occur in response to less initial temperatures stress for more commonly-experienced thermal extremes; low temperatures for *A. picea*, high temperatures for *A. carolinensis*. Second, the *genetic assimiliation* hypothesis [@waddington1953; @sikkink2014] proposes that exposure to more frequent stressors selects for a shift from inducible to constitutive expression of stress-response genes. This hypothesis predicts that transcripts responsive to high temperatures in *A. picea* will have higher constitutive expression in *A. carolinensis*, while transcripts responsive to low temperatures in *A. carolinensis* will have higher constitutive expression in *A. picea*. Finally, the *tolerance* hypothesis [@cowles1939; @fields2001] proposes that existing inducible stress responses become insufficient or prohibitively costly as environmental stressors increase in frequency, resulting in a shift away from induced response in favor of structural [@lockwood2012] or behavioral [@gordon2013; @sunday2014] adapations. This hypothesis predicts adaptation to stress should be associated with lower transcriptional responsiveness and less sensitivity to temperature perturbation, as well as a shift to an alternate suite of tolerance genes and pathways. Of course, these hypotheses are not mutually exclusive. By taking a reactionome approach, we can quantify the relative importance of each of these mechanisms to thermal adapatation as the proportion of the transcriptome matching each of these predicted patterns.

@@

The potential for many species to persist in face of contemporary climate change will depend in part upon their thermal tolerance. However, for most species little is known about the plasticity or adaptive changes in gene expression that underlie thermal tolerance. By using the novel *reactionome* approach, we were able to quantitatively describe changes in transcript expression across a thermal gradient. In particular, by characterizing the thermal reactionomes of the ant species *Aphaenogaster picea* and *A. carolinensis* we found patterns consistent with different mechanisms of thermal tolerance in these two species. Overall, we found that about 8% of transcripts (~9,000) are thermally responsive in each species. About 70% of these transcripts have a non-linear response to temperature that differs between species, which would not have been detected by standard differential expression approaches.

Under the *enhanced defense* hypothesis, stress-adapted species are hypothesized to induce a stronger and earlier response to extreme conditions. We found evidence for this hypothesis from the cool-climate species *A. picea* which induced more transcripts in response to low temperatures than did *A. carolinensis* (Table 1, Fig. 1). However, *A. picea* also induced more transcripts in response to high temperatures than did *A. carolinensis*. We also found that higher temperatures were required to induce the expression response for *A. carolinensis* than *A. picea* towards the high thermal limit (Fig. 2B, D), consistent with the *tolerance* but not the *enhanced defense* hypothesis for *A. carolinensis*. There was no difference between species in temperature of response to low temperatures (Fig. 2A,C). ...compare to Drosophila...

Exposure to more frequent stressors is hypothesized to select for a shift from inducible to constitutive expression of stress-response transcripts under the *genetic assimilation* hypothesis. In contrast to this prediction, constitutive expression levels at 19.5°C were higher in *A. carolinensis* than *A. picea* in nearly all comparisons. Consistent with our results but at a much shorter time-scale, Sikkink et al. [-@sikkink2014] found no evidence for genetic assimilation at the expression level after 10 generations of selection for heat tolerance in *Caenorhabditis remanei*, even though there was a substantial increase in heat tolerance and loss of plasticity in the evolved populations. While our results did not the genetic assimilation hypothesis, the observation of higher constitutive expression of thermally-responsive transcripts is again consistent with the *tolerance* hypothesis for *A. carolinensis*.

The exception the higher constitutive expression of *A. carolinensis* relative to *A. picea* transcripts was for transcripts that had **High** or **Bimodal** expression in *A. picea* and were not responsive in *A. carolinensis*. Rather than being constitutively elevated as predicted, these transcripts were not expressed (...mean TPM...) in *A. carolinensis*. This further supports the *enhanced defense* hypothesis for adaptation of *A. picea* to cold temperatures as these are novel genes recruited to the low temperature response in this species.  ...gene ontology terms...

Taken together, these results indicate that these two species have distinct mechanisms of response to thermal extremes. *Aphaenogaster picea* mounts a classical cellular stress response [@kultz2005; @richter2010] to resist the impacts of thermal stress, requiring increased metabolism, while *A. carolinensis* relies upon constitutive defenses [@barshis2013], proteome stability [@ghosh2010] or behavior [@sunday2014] to tolerate thermal stress.  If this hypothesis is correct, we would expect that candidate genes such as heat shock proteins would be over-represented in the thermally-responsive category for *A. picea*. Consistent with this, of the `r nrow(hsp_all)` transcripts annotated to heat shock proteins in the transcriptome, `r length(which(hsp_responsive$A22.type %in% c("Bimodal", "High", "Low")))` were upregulated to thermal extremes in *A. picea* while only `r length(which(hsp_responsive$Ar.type %in% c("Bimodal", "High", "Low")))` were upregulated in *A. carolinensis*. In contrast, by shutting-down molecular activity, must rely upon other mechanisms such as behavior or increased proteome stability to tolerate thermal stress. These differences are in line with the expectations if *A. picea*, a northern species with fewer good foraging days, must forage in suboptimal conditions whereas *A. carolinensis* can, with a longer season, afford to restrain from foraging in suboptimal conditions. In fact, restrained foraging under stressful conditions by the red harvester ant *Pogonomyrmex barbatus* has been shown to increase colony fitness [@gordon2013].

Critical maximum and minimum temperatures are hypothesized to be genetically correlated [@addo-bediako2000; @hoffmann2013], but this was not evident in terms of gene expression.  Only 15-20% of transcripts upregulated in response to temp are bimodal (i.e. generalized stress response), and although upper activation thresholds differed significantly between species, lower thresholds were virtually identical. Implications ...

@@

A major contribution of this study is the reaction norm approach for gene expression data (but see Hodgins-Davis et al [-@hodgins-davis2012] for a similar approach using differential expression among many levels). A number of issues have hampered RNA-seq studies to date. Namely, lists of differentially expressed transcripts are prone to false positives [@gonzalez2013], depend on the genetic background of the organism [@sarup2011], identify transcripts that are correlated, but not necessarily causative, with phenotypes [@feder2005] and are prone to "storytelling" [@pavlidis2012]. Our findings are robust to these issues as we focus on the average change in the shape of the reaction norms across many hundreds of responsive transcripts in each species. Thus, while we use gene ontology information to explain our main results, the key findings about the number of responsive transcripts and shifts in temperatures of induction do not depend on functional annotation. Similarly, in the plant and animal breeding literature there has been a shift from using genome-wide association methods which are prone to the same problems listed above, to using genomic selection methods which provide a direct means to predict a population's response to selection [@goddard2007].

@@

In this work, we have brought reaction norms to the genomic era by characterizing the thermal reactionomes of two temperate ant species, *Aphaenogaster picea* and *A. carolinensis*. At least 10% of their transcriptomes are thermally responsive. For both species, more transcripts responded to low than high temperatures. Together with physiological data showing more variation in critical minimum than maximum temperatures, characterization of the reactionome suggests that these species, and likely other ectotherms, have less adaptive potential in response to high than low temperatures. Our results also indicate that these two ant species have different responses to thermal extremes. *A. picea* responds by increasing metabolism and activating heat shock proteins and other protective molecules, whereas *A. carolinensis* shuts down metabolism and must rely on other mechanisms for thermal tolerance. The thermal reactionomes of these two species provide key insights into the genetic basis of thermal tolerance, and a resource for the future study of ecological adaptation in ant species. Finally, the reactionome itself illustrates a new direction for characterizing acclimation and adaptation in a changing climate.

@@

To describe the overall molecular toolkit for thermal tolerance in *Aphaenogaster*, we performed gene set enrichment analysis (GSEA) on the intersection of each expression category in the two species using the `parentChild` algorithm [@grossmann2007] from the package `topGO` [@alexa2006] in R [@r_core_team2013]. Briefly, this approach identifies GO terms that are overrepresented in the significant transcripts relative to all GO terms in the transcriptome, accounting for dependencies among the GO terms. We clustered enriched GO terms from the `topGO` analysis by semantic similarity using the package `GOSemSim` [@yu2010] to reduce redundancy.

@@

The raw Illumina reads are available from the NCBI Short Read Archive .... The script for analysis and version history is available at (http://github.com/johnstantongeddes/ApTranscriptome).

### 2014-10-10 doc/PNAS_cover_letter.tex
edits to text, mostly Discussion

First, we make a methodological contribution by extending the reaction
norm approach to the study of gene expression. We define the
\emph{reactionome} as a characterization of the reaction norm for all
genes in an organism's transcriptome across an environmental gradient. 
We believe that this approach is an important contribution as it allows us
to make quantitative comparisons of plasticity in gene expression between
species. Importantly, this analysis does not depend on the functional 
annotation of the genes, and thus is robust to criticisms of "storytelling" 
in RNAseq studies based on annotation of differentially expressed 
genes (Pavlidis et al. 2012 MBE).

Second, by applying the reactionome approach to two related ant species,
\emph{Aphaenogaster picea} and \emph{A. carolinensis}, from colder and
warmer climates, respectively, we gain insight on the relationship
between short-term physiological acclimation and longer-term adaptive
change in response to thermal extremes. In particular, we find that
adaptation to chronic high temperatures involves different mechanisms
than those used to cope with rarely encountered stressors, which may
limit adaptive potential of northern species in the face of rapidly
shifting climatices.

@@

\textbf{Suggested NAS Members:} James H. Brown, Bert Hoelldobler, Trudy F. Mackay, Nevo Eviata, Johanna Schmitt

\textbf{Suggested reviewers:}

\begin{itemize}
  \item{Dr. Nicholas Teets, Postdoc, University of Florida}
  \item{Dr. Vanessa Kellerman, Postdoc, Monash University}
  \item{Dr. Pavlos Pavlidis, Postdoc, Heidelberg Institute of Theoretical Studies}
  \item{Dr. Stephen Palumbi, Professor, Hopkins Marine Station, Stanford University}
  \item{Dr. Joel Kingsolver, Professor, University of North Carolina}
\end{itemize}

\noindent{Thank you for consideration of our work,} \\
\includegraphics[scale=1]{JSG_signature.png}  \\
Dr. John Stanton-Geddes

>>>>

In face of contemporary climate change, many species are predicted to be at 
risk of extinction. However, there is also evidence that species can physiologically
acclimate and rapidly adapt to changing environmental conditions, potentially
rescuing them from extinction. This paradox poses a dilemma for evolutionary
ecologists. In this study, we evaluate plasticity and adaptive differentiation
in gene expression of closely related cool and warm-climate ant species.

We present a novel analysis by extending the reaction
norm approach to the study of gene expression. We define the
\emph{reactionome} as a characterization of the reaction norm for all
genes in an organism's transcriptome across an environmental gradient. 
We believe that this approach is an important contribution as it allows us
to make quantitative comparisons of plasticity in gene expression between
species. Importantly, this analysis does not depend on the functional 
annotation of the genes, and thus is robust to criticisms of "storytelling" 
in RNAseq studies based on annotation of differentially expressed 
genes (Pavlidis et al. 2012 MBE).

By applying the reactionome approach to two related ant species,
\emph{Aphaenogaster picea} and \emph{A. carolinensis}, from colder and
warmer climates, respectively, we gain insight on the relationship
between short-term physiological acclimation and longer-term adaptive
change in response to thermal extremes. In particular, we find that
adaptation to chronic high temperatures involves different mechanisms
than those used to cope with rarely encountered stressors, which may
limit adaptive potential of northern species in the face of rapidly
shifting climatices.

@@

\textbf{Suggested NAS Members:} Trudy F. Mackay, Nevo Eviata, George N. Somero

\textbf{Suggested reviewers:}  Dr. Nicholas Teets, Postdoc, University of Florida; Dr. Vanessa Kellerman, Postdoc, Monash University; Dr. Pavlos Pavlidis, Postdoc, Heidelberg Institute of Theoretical Studies; Dr. Stephen Palumbi, Professor, Hopkins Marine Station, Stanford University; Dr. Joel Kingsolver, Professor, University of North Carolina

\noindent{Thank you for consideration of our work,} \\
\includegraphics[scale=1]{JSG_signature.png}  \\

Dr. John Stanton-Geddes
